---
title: "Three Age-Group Analysis"
author: "Jiajun"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    css: ["css/text_style.css", "css/table_style.css"]
    theme: sandstone
    toc_depth: 3
    highlight: tango
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 10, fig.height = 5)
```

## Library and Data

```{r load-library-and-data, class.source='fold-show'}
library(survival)
library(tidyverse)
library(penalized)
library(kableExtra)
library(stringr)
library(patchwork)

source("../R/utils.R")

library(readxl)
raw_data <- read_excel("../data/Aim3-early-adversity-subjects-F-survival-2021-01-07.xlsx")
```

Raw data here refers to the larger dataset.

## EDA

Below is the Kaplan-Meier estimation of the survival function on the raw data.

```{r eda}
km <- survfit(Surv(raw_data$age, raw_data$dead) ~ 1, conf.type = "log-log")
ggplot() +
  geom_line(aes(x = km$time, y = km$surv)) +
  geom_vline(aes(xintercept = 1), color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = 5), color = "red", linetype = "dashed") +
  labs(x = expression("Time (year)"), y = expression("Survival"),
       title = "Kaplan-Meier estimate") +
  scale_x_continuous(breaks = seq(0, 25, 1)) +
  theme_classic()
```

## Analysis Process

We conduct four models on three datasets derived from the raw data.

Four models:

- Cox model of survival data versus all the main effects (six adversities)
- Cox model of survival data versus main effects and interaction terms selected by lasso
- Cox model of survival data versus main effects and interaction terms selected by stepwise BIC
- Cox model of survival data versus `adv_mom`, sum of other adversities `adv_other_cumulative` and their interaction term

Three datasets:

- `age_below_1`: Include all the samples. For those whose `age` > 1, use `age` = 1, and `dead` = FALSE
- `age_above_1_below_5`: Include samples of `age` > 1. For those whose `age` > 5, use `age` = 5, and `dead` = FALSE
- `age_above_5`: Include samples of `age` > 5.

Note that for the `age_below_1` dataset, we exclude `adv_sib` from all models.

```{r create-dataset}
age_below_1 <- raw_data %>%
  select(starts_with("adv_"), age, dead, -adv_cumulative) %>%
  mutate(dead = if_else(age > 1, FALSE, dead)) %>%
  mutate(age = if_else(age > 1, 1, age))

age_above_1_below_5 <- raw_data %>%
  filter(age >= 1) %>%
  select(starts_with("adv_"), age, dead, -adv_cumulative) %>%
  mutate(dead = if_else(age > 5, FALSE, dead)) %>%
  mutate(age = if_else(age > 5, 5, age))

age_above_5 <- raw_data %>%
  filter(age >= 5) %>%
  select(starts_with("adv_"), age, dead, -adv_cumulative)
```

Below is the summary of three datasets. The number of events refers to the number of uncensored death.


```{r summary-of-three-datasets}
data.frame(
  nobs = c(nrow(age_below_1), nrow(age_above_1_below_5), nrow(age_above_5)),
  nevt = c(sum(age_below_1$dead), sum(age_above_1_below_5$dead), sum(age_above_5$dead))
) %>%
  `row.names<-`(c("Age 1-", "Age 1+, 5-", "Age 5+")) %>%
  kable(row.names = TRUE,
        col.names = c("Number of Samples", "Number of Events")) %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 18)
```


## Cox Model - Main Effects {.tabset .tabset-fade .tabset-pills}

```{r full-model-helper}
run_full_model <- function(data, below_one = FALSE) {
  varnames = names(data)[startsWith(names(data), "adv_")]
  if (below_one) {
    varnames = varnames[varnames != "adv_sib"]
  }
  full_formula = paste("Surv(age, dead) ~", paste0(varnames, collapse = "+ "))
  full_model = coxph(
    as.formula(full_formula),
    data = data
  )
  return (full_model)
}
```


### Age below 1

```{r age-below-1-full-model}
full_age_below_1 <- run_full_model(age_below_1, below_one = TRUE)
summary(full_age_below_1)
```



### Age above 1 and below 5

```{r age-above-1-below-5-full-model}
full_age_above_1_below_5 <- run_full_model(age_above_1_below_5)
summary(full_age_above_1_below_5)
```

### Age above 5

```{r age-above-5-full-model}
full_age_above_5 <- run_full_model(age_above_5)
summary(full_age_above_5)
```


## {-}

```{r summary-of-full-model, eval=FALSE, include=FALSE}
sig_vars <- sapply(
  list(full_age_below_1, full_age_above_1_below_5, full_age_above_5),
  function(cox_obj, level=0.05) {
    varnames = names(cox_obj$coef)
    pvals = summary(cox_obj)$coef[, "Pr(>|z|)"]
    sig_vars = varnames[abs(pvals) < level]
    sig_pvals = pvals[abs(pvals) < level]
    sig_codes = symnum(sig_pvals, na = FALSE,
                       cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
                       symbols = c("***", "**", "*", ".", " "))
    return (paste(sig_vars, sig_codes))
  }
)

max_length <- max(sapply(sig_vars, length))
sig_vars <- lapply(sig_vars, function(v) c(v, rep(NA, max_length - length(v))))

options(knitr.kable.NA = '')
data.frame(do.call(rbind, sig_vars)) %>%
  `row.names<-`(c("Age 1-", "Age 1+, 5-", "Age 5+")) %>%
  kable(col.names = c("", "Signif. Variables")) %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 18)
```


```{r comparison-plot-full-model, eval=FALSE, include=FALSE}
p1 <- plot_forest(full_age_below_1)
p2 <- plot_forest(full_age_above_1_below_5)
p3 <- plot_forest(full_age_above_5)

p1 + p2 + p3
```



## Cox Model - Lasso {.tabset .tabset-fade .tabset-pills}

```{r lasso-helper}
get_penalized_covariates <- function(data, below_one=FALSE) {
  data = select(data, starts_with("adv_"))
  if (below_one) {
    data = select(data, -adv_sib)
  }
  penalized_covariates <- model.matrix(
    as.formula("~ . + .:."),
    data = data.frame(data)
  )
  penalized_covariates <- penalized_covariates[, colnames(penalized_covariates) != "(Intercept)"]
  return (penalized_covariates)
}

get_opt_lambda_of_lasso <- function(data, penalized_covariates, seed = 42, fold = 5) {
  set.seed(seed)
  opt <- optL1(
    Surv(data$age, data$dead),
    penalized = penalized_covariates,
    standardize = T,
    fold = fold,
    trace = F
  )
  return (opt$lambda)
}

get_nonzero_varnames_of_opt_lasso <- function(data, penalized_covariates, opt_lambda) {
  final = penalized(
    Surv(data$age, data$dead),
    penalized = penalized_covariates,
    standardize = T,
    lambda1 = opt_lambda,
    trace = F
  )
  return (names(final@penalized)[final@penalized != 0])
}

get_varnames_in_final_model <- function(nonzero_varnames) {
  # append main effects select by interaction terms
  ans = c()
  for (varname in nonzero_varnames) {
    if (str_detect(varname, ":")) {
      ans = c(ans, str_split(varname, ":")[[1]])
    }
    ans = c(ans, varname)
  }
  return (unique(ans))
}

run_cox <- function(data, varnames, penalized_covariates) {
  formula = paste("Surv(age, dead) ~", paste0(varnames, collapse = "+ "))
  cox <- coxph(
    as.formula(formula),
    data = data.frame(
      cbind(penalized_covariates,
            select(data, c("age", "dead"))
      ))
  )
  return (cox)
}
```

### Age below 5

```{r lasso-age-below-1}
penalized_below_1 <- get_penalized_covariates(age_below_1, below_one = TRUE)
opt_lambda_below_1 <- get_opt_lambda_of_lasso(
  data = age_below_1,
  penalized_covariates = penalized_below_1
)

nonzero_vars_below_1 <- get_nonzero_varnames_of_opt_lasso(
  data = age_below_1,
  penalized_covariates = penalized_below_1,
  opt_lambda = opt_lambda_below_1
)

vars_below_1 <- get_varnames_in_final_model(nonzero_vars_below_1)
cox_below_1 <- run_cox(
  data = age_below_1,
  varnames = vars_below_1,
  penalized_covariates = penalized_below_1
)

summary(cox_below_1)
```


### Age above 1 and below 5

```{r lasso-age-above-1-below-5}
penalized_above_1_below_5 <- get_penalized_covariates(age_above_1_below_5)
opt_lambda_above_1_below_5 <- get_opt_lambda_of_lasso(
  data = age_above_1_below_5,
  penalized_covariates = penalized_above_1_below_5
)
nonzero_vars_above_1_below_5 <- get_nonzero_varnames_of_opt_lasso(
  data = age_above_1_below_5,
  penalized_covariates = penalized_above_1_below_5,
  opt_lambda = opt_lambda_above_1_below_5
)

vars_above_1_below_5 <- get_varnames_in_final_model(nonzero_vars_above_1_below_5)
cox_above_1_below_5 <- run_cox(
  data = age_above_1_below_5,
  varnames = vars_above_1_below_5,
  penalized_covariates = penalized_above_1_below_5
)

summary(cox_above_1_below_5)
```


### Age above 5

```{r lasso-age-above-5}
penalized_above_5 <- get_penalized_covariates(age_above_5)
opt_lambda_above_5 <- get_opt_lambda_of_lasso(
  data = age_above_5,
  penalized_covariates = penalized_above_5
)
nonzero_vars_above_5 <- get_nonzero_varnames_of_opt_lasso(
  data = age_above_5,
  penalized_covariates = penalized_above_5,
  opt_lambda = opt_lambda_above_5
)

vars_above_5 <- get_varnames_in_final_model(nonzero_vars_above_5)
cox_above_5 <- run_cox(
  data = age_above_5,
  varnames = vars_above_5,
  penalized_covariates = penalized_above_5
)

summary(cox_above_5)
```



## {-}

```{r summary-of-model-by-lasso, eval=FALSE, include=FALSE}
lasso_sig_vars <- sapply(
  list(cox_below_1, cox_above_1_below_5, cox_above_5),
  function(cox_obj, level=0.05) {
    varnames = names(cox_obj$coef)
    pvals = summary(cox_obj)$coef[, "Pr(>|z|)"]
    sig_vars = varnames[abs(pvals) < level]
    sig_pvals = pvals[abs(pvals) < level]
    sig_codes = symnum(sig_pvals, na = FALSE,
                       cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
                       symbols = c("***", "**", "*", ".", " "))
    return (paste(sig_vars, sig_codes))
  }
)

data.frame(t(lasso_sig_vars)) %>%
  `row.names<-`(c("Age 1-", "Age 1+, 5-", "Age 5+")) %>%
  kable(col.names = c("", "Signif. Variables")) %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 18)
```


```{r comparison-plot-lasso-model, eval=FALSE, include=FALSE}
p1 <- plot_forest(cox_below_1)
p2 <- plot_forest(cox_above_1_below_5)
p3 <- plot_forest(cox_above_5)

p1 + p2 + p3
```


## Cox Model - Stepwise BIC {.tabset .tabset-fade .tabset-pills}

### Age below 1

```{r cox-model-bic-age-below-1}
full_bic_below_1 <- coxph(
  as.formula(Surv(age, dead) ~ . + .:.),
  data = select(age_below_1, -adv_sib)
)

step_bic_below_1 <- step(
  full_bic_below_1,
  k = log(nrow(age_below_1)),
  #scope = list(lower= ~ adv_density + adv_rain + adv_mom + adv_mom_rank + 
  #  adv_mom_dsi),
  trace = F
)

summary(step_bic_below_1)
```

### Age above 1, below 5

```{r cox-model-bic-age-above-1-below-5}
full_bic_above_1_below_5 <- coxph(
  as.formula(Surv(age, dead) ~ . + .:.),
  data = age_above_1_below_5
)

step_bic_above_1_below_5 <- step(
  full_bic_above_1_below_5,
  k = log(nrow(age_above_1_below_5)),
  #scope = list(lower= ~ adv_density + adv_rain + adv_mom + adv_mom_rank + 
  #  adv_mom_dsi + adv_sib),
  trace = F
)

summary(step_bic_above_1_below_5)
```


### Age above 5

```{r cox-model-bic-age-above-5}
full_bic_above_5 <- coxph(
  as.formula(Surv(age, dead) ~ . + .:.),
  data = age_above_5
)

step_bic_above_5 <- step(
  full_bic_above_5,
  k = log(nrow(age_above_5)),
  #scope = list(lower= ~ adv_density + adv_rain + adv_mom + adv_mom_rank + 
  #  adv_mom_dsi + adv_sib),
  trace = F
)

summary(step_bic_above_5)
```


## {-}

```{r summary-of-model-by-bic, eval=FALSE, include=FALSE}
bic_sig_vars <- sapply(
  list(step_bic_below_1, step_bic_above_1_below_5, step_bic_above_5),
  function(cox_obj, level=0.05) {
    if (is.null(cox_obj$coef)) {
      return ("Null model")
    }
    varnames = names(cox_obj$coef)
    pvals = summary(cox_obj)$coef[, "Pr(>|z|)"]
    sig_vars = varnames[abs(pvals) < level]
    sig_pvals = pvals[abs(pvals) < level]
    sig_codes = symnum(sig_pvals, na = FALSE,
                       cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
                       symbols = c("***", "**", "*", ".", " "))
    return (paste(sig_vars, sig_codes))
  }
)
max_length <- max(sapply(bic_sig_vars, length))
bic_sig_vars <- lapply(bic_sig_vars, function(v) c(v, rep(NA, max_length - length(v))))

options(knitr.kable.NA = '')
data.frame(do.call(rbind, bic_sig_vars)) %>%
  `row.names<-`(c("Age 1-", "Age 1+, 5-", "Age 5+")) %>%
  kable(col.names = c("", "Signif. Variables")) %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 18)
```


```{r comparison-plot-bic-model, eval=FALSE, include=FALSE}
p1 <- plot_forest(step_bic_below_1)
p2 <- plot_forest(step_bic_above_1_below_5)
p3 <- plot_forest(step_bic_above_5)

p2 + p3
```



## Cox model - Cumulative {.tabset .tabset-fade .tabset-pills}


```{r cox-model-of-cumulative-adversities-helper-function}
run_cox_with_cumulative <- function(data) {
  data = data %>%
    mutate(adv_other_cumulative = adv_sib+adv_rain+adv_mom_dsi+adv_mom_rank+adv_density) %>%
    select(age, dead, adv_mom, adv_other_cumulative)
  cox = coxph(
    formula = as.formula("Surv(age, dead) ~ . + adv_mom:adv_other_cumulative"),
    data = data
  )
  return (cox)
}
```

### Age below 1

```{r cumulative-age-below-1}
cox_below_1 <- run_cox_with_cumulative(age_below_1)
summary(cox_below_1)
```

### Age above 1 and below 5

```{r cumulative-age-above-1-below-5}
cox_above_1_below_5 <- run_cox_with_cumulative(age_above_1_below_5)
summary(cox_above_1_below_5)
```

### Age above 5

```{r cumulative-age-above-5}
cox_above_5 <- run_cox_with_cumulative(age_above_5)
summary(cox_above_5)
```


## {-}

```{r summary-of-model-with-cumulative, eval=FALSE, include=FALSE}
cumulative_sig_vars <- sapply(
  list(cox_below_1, cox_above_1_below_5, cox_above_5),
  function(cox_obj, level=0.05) {
    varnames = names(cox_obj$coef)
    pvals = summary(cox_obj)$coef[, "Pr(>|z|)"]
    sig_vars = varnames[abs(pvals) < level]
    sig_pvals = pvals[abs(pvals) < level]
    sig_codes = symnum(sig_pvals, na = FALSE,
                       cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
                       symbols = c("***", "**", "*", ".", " "))
    return (paste(sig_vars, sig_codes))
  }
)

max_length <- max(sapply(cumulative_sig_vars, length))
cumulative_sig_vars <- lapply(cumulative_sig_vars, function(v) c(v, rep(NA, max_length - length(v))))

options(knitr.kable.NA = '')
data.frame(do.call(rbind, cumulative_sig_vars)) %>%
  `row.names<-`(c("Age 1-", "Age 1+, 5-", "Age 5+")) %>%
  kable(col.names = c("", "Signif. Variables")) %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 18)
```



```{r comparison-plot-cumulative-model, eval=FALSE, include=FALSE}
p1 <- plot_forest(cox_below_1)
p2 <- plot_forest(cox_above_1_below_5)
p3 <- plot_forest(cox_above_5)

p1 + p2 + p3
```



## Discussion


- For the full model with main effects of all six adversities, group `age_below_1` and `age_above_5` behave similarly in terms of the significant variable `adv_mom`. Group `age_above_1_below_5` has one more significant relationship with `adv_dsi`, and the significance level of `adv_mom` is much lower, indicating the hazard might work differently within this age range.

- For the model selected by lasso, the resulting model picks quite a small fraction of variables, with no interaction terms in the `age_below_1` group, and one non-significant interaction terms in the `age_above_1_below_5` group. Interaction term `adv_mom:adv_rain` is significant in the `age_above_5` group while the other interaction term `adv_mom:adv_sib` is picked but not significant.

- For the model selected by stepwise BIC, no interaction terms are selected in `age_below_1` and `age_above_1_below_5`, which is the same result as in models selected by lasso. Meanwhile, `adv_mom:adv_rain` is picked and significant in `age_above_5`.

- Interestingly, the cumulative model in the "age_below_1" group shows significance for the cumulative variable and the interaction term. There does not seem to be much worth noting in the other two groups.




