---
title: "Lasso and Cox Regression"
author: "Jiajun"
date: "1/2/2020"
output:
  rmdformats::readthedown:
    css: ["css/text_style.css", "css/table_style.css"]
    theme: sandstone
    toc_depth: 3
    highlight: tango
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Library and Data

```{r load-library-and-data, class.source='fold-show'}
library(survival)
library(tidyverse)
library(penalized)
library(latex2exp)

library(readxl)
raw_data <- read_excel("../data/Aim3-early-adversity-subjects-F-survival-2020-10-09.xlsx")
```

## EDA

```{r eda-check-na-and-data-types}
# sum(is.na(raw_data))

data <- raw_data %>%
  mutate(sex = factor(sex),
         matgrp = factor(matgrp),
         status = factor(status))

summary(data)
```

- No `NA` values
- Response (survival time): `age`
- Categorical variables (3):
  1. `sex`: only contains Female
  2. `matgrp`: social group of mothers
  3. `status`: 0, alive; 1, dead; 3, dropped out due to observer events
- Logical variables (6): `dead`, `adv_density`, `adv_rain`, `adv_mom`, `adv_sib`, `adv_mom_rank`, `adv_mom_dsi`
- Date (2): `birth`, `statdate`
- Numerical variables (1): `adv_cumulative`
- Other (1): `sname`


`sex` is the same for all samples so there is no need to include it. Here I would treat the `status` = 3 as non-informative censoring so that `dead` is included as censoring variable. The date does not seem useful at this stage since I am not going to consider temporal information. Also, `sname` is not informative as raw string type.

Hence, the censoring variable is `dead` and the predictors I use are

- six adversities, `adv_cumulative`
- interactions: `adv_mom`:`other adversity`, `adv_rain`:`other adversity`


```{r prepare-penalized-covariates}
data_sub <- data %>% select(starts_with("adv_"))

# interactions selected are between adv_mom:other, and adv_rain:other
formula <- "~ . + adv_mom:(.-adv_cumulative) + adv_rain:(.-adv_cumulative)"
penalized_covariates <- model.matrix(as.formula(formula) , data = data.frame(data_sub))

# remove intercept ready for penalized covariates
penalized_covariates <- penalized_covariates[, colnames(penalized_covariates) != "(Intercept)"]
```


## Lasso

```{r lasso-methods-for-variable-selection}
set.seed(42)

cv <- profL1(
  Surv(data$age, data$dead),
  penalized = penalized_covariates,
  standardize = T,
  fold = 5,
  minlambda1 = 5,
  maxlambda1 = 25,
  trace = F
)

set.seed(42)
opt <- optL1(
  Surv(data$age, data$dead),
  penalized = penalized_covariates,
  standardize = T,
  fold = 5,
  trace = F
)
```


```{r plot-cv-results, echo=FALSE}
ggplot() +
  geom_line(aes(x = cv$lambda, y = cv$cvl)) +
  geom_vline(xintercept = opt$lambda,
             color = "gray", size = 2) +
  labs(x = TeX("$\\lambda$"), y = "Partial Likelihood",
       caption = "Cross-validated partial likelihood for a range values of lambda. The vertical gray line shows the global maximum.") +
  theme_bw()
```


## Stepwise Lasso Fit

```{r stepwise-lasso-fit, fig.height=6, fig.width=12}
stepwise <- penalized(
  Surv(data$age, data$dead),
  penalized = penalized_covariates,
  standardize = T,
  steps = 20,
  lambda1 = 1,
  trace = F
)

plotpath(
  stepwise,
  labelsize = 0.8,
  standardize = T
)

abline(
  v = opt$lambda,
  col = "gray",
  lwd = 2
)
```



## Nonzero covariates

```{r coef}
final <- penalized(
  Surv(data$age, data$dead),
  penalized = penalized_covariates,
  standardize = T,
  lambda1 = opt$lambda,
  trace = F
)
nonzero_varnames <- names(final@penalized)[final@penalized != 0]
nonzero_varnames
```

Variable selection by lasso picks out 5 covariates: `adv_density`, `adv_mom`, `adv_rain:adv_mom`, `adv_mom:adv_sib`, and `adv_mom:adv_mom_ds`.


## Cox Regression

```{r cox-model-with-selected-covariates}

formula = paste("Surv(data$age, data$dead) ~", paste0(nonzero_varnames, collapse = "+ "))

cox <- coxph(
  as.formula(formula),
  data = data.frame(penalized_covariates)
)

summary(cox)
```


## Rerun with composite adversity

""LASSO pick up five terms (two main effects and three interactions (all with maternal loss), but none of the interaction term is significant.I was wondering if you can create a composite variable of whether there is one adversity besides mom death (sum of all other adversities>0), and rerun the analysis with the two main effects and the interaction between mom death and this composite adversity.""

```{r rerun-with-modified-covariates}
data_composite <- raw_data %>%
  mutate(adv_composite = adv_rain+adv_sib+adv_mom_dsi) %>%
  select(adv_density, adv_mom, adv_composite)

cox_composite <- coxph(
  Surv(data$age, data$dead) ~ adv_density + adv_mom + adv_mom*adv_composite,
  data = data_composite
)

summary(cox_composite)
```




